---
title: "COVID-19 EDA"
author:
- Diego G. Davila
- Margaret Gardner
- Joelle Bagautdinova
date: 'Due before midnight, May 1st'
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
  pdf_document:
    toc_depth: '4'
    number_sections: yes
urlcolor: blue
---

# COVID-19 Datasets EDA

```{r Setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output

# Package setup
if(!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, dplyr, ggplot2, ggthemes, data.table, lubridate,
               GGally, RColorBrewer, ggsci, plotROC, usmap,
               plotly, ggpubr, vistime, skimr, glmnet, dgCMatrix, leaps, car)
```

**Overall goal:**

In this EDA, we are interested in combining a worldwide daily COVID cases dataset with a vaccination and a COVID variants dataset to follow the worldwide dynamics of virus spread, variant emergence/spread and the effect of vaccination in mitigating the number of cases/deaths.These datasets were collected from the following sources:

* [COVID vaccines](https://www.kaggle.com/datasets/gpreda/covid-world-vaccination-progress)
* [COVID cases](https://www.kaggle.com/datasets/josephassaker/covid19-global-dataset)
* [COVID variants](https://www.kaggle.com/datasets/gpreda/covid19-variants)


**Relevant questions:** 

* How did COVID-19 spread since the beginning of the pandemic (worldwide)?
* How is vaccination affecting virus spread? 
* Which country is using which vaccine?
* In which country the vaccination program is more advanced?
* Which country has higher daily vaccination rates (i.e. proportionally to its population)? Does this affect the number of cases/deaths?
* What variants are sequenced and in which country?
* Where did variants originate from and how did they spread over time?
* Which regions show high vs low variant diversity (i.e. a dominant variant is proliferating)?
* Does the type of dominant variant (e.g., delta, omicron) influence the efficicay of vaccines/number of cases/number of deaths?


## Numerical EDA

```{r}

# load data
df_cases <- read.csv("data/worldometer_coronavirus_daily_data.csv")

# clean data
df_cases <- df_cases %>%
  mutate(country = as.factor(country),
         date = lubridate::mdy(date),
         year = lubridate::year(date),
         month_num = lubridate::month(date),
         week = as.factor(lubridate::week(date))
  )

df_cases$month <- month.abb[df_cases$month_num] # convert to month names
skim(df_cases)

# create new monthly and weekly cases/deaths variables
df_cases <- df_cases %>%
  # monthly
  group_by(month_num, year, country) %>% 
  mutate(
    monthly_new_cases = sum(daily_new_cases, na.rm = TRUE),
    monthly_new_deaths = sum(daily_new_deaths, na.rm = TRUE),
    month_year = make_date(year, month_num) 
  ) %>%
  ungroup() %>%
  # weekly
  group_by(week, year, country, month_num) %>% 
  mutate(
    weekly_new_cases = mean(daily_new_cases, na.rm = TRUE),
    weekly_new_deaths = mean(daily_new_deaths, na.rm = TRUE),
    month_year = make_date(year, month_num)
    )


df_vac <- read.csv("data/country_vaccinations.csv")
# clean data
df_vac <- df_vac %>%
  mutate(country = as.factor(country),
         country = recode_factor(country, `United States` = "USA"),
         iso_code = as.factor(iso_code),
         vaccines = as.factor(vaccines),
         date = lubridate::ymd(date),
         year = lubridate::year(date),
         month_num = lubridate::month(date))

df_vac$month <- month.abb[df_vac$month_num] # convert to month names

skim(df_vac)

# df_variants <- read.csv("data/covid-variants.csv")
# # clean data
# df_variants <- df_variants %>%
#   mutate(country = as.factor(location),
#          variant = as.factor(variant))  %>%
#   select(-location)
# 
# skim(df_variants)

```

* COVID cases: 
  + all variables of interest have at least an 84% completion rate, so all variables are usable. 
  + there are 225 unique countries listed in this dataset. 
* Vaccines: 
  + some variables such as `total_vaccinations` have low completion rates (<50%) as these are reported every other week/once a month in some countries. 

## Visual EDA


### COVID-19 cases 2020-2022

Let's first look at the evolution of COVID-19 cases from the start of the pandemic (March 2020) until the last census (March 2022).



```{r}
# let's see how the total cumulative covid cases evolved by country
plot_cum_cases <- df_cases %>%
  # filter(country %in% c("USA", "China", "India")) %>% # to show only some countries
  # group_by(month, year) %>% 
  # filter(date == min(date)) %>% #  to show monthly evolution (curves are slightly less smooth)
  ggplot(aes(x = date, y = cumulative_total_cases, color = country, group = country)) + 
    geom_line(show.legend = FALSE) + 
    theme_bw() +
    ggtitle("Worldwide cumulative daily new cases") +
    theme(legend.position="bottom")

ggplotly(plot_cum_cases)

# TODO
# world map of cumulative cases

```

What about a monthly evolution? Let's see what that looks like: 

```{r}

# what about the number of new cases?
plot_cases_monthly <- df_cases %>%
  ggplot(aes(x = month_year, y = monthly_new_cases, color = country, group = country)) + 
    geom_line(show.legend = FALSE) + 
    theme_bw() +
    ggtitle("Worldwide monthly new cases") +
    theme(legend.position="bottom")

ggplotly(plot_cases_monthly)



# and the number of new deaths?
plot_deaths_monthly <- df_cases %>%
  ggplot(aes(x = month_year, y = monthly_new_deaths, color = country, group = country)) + 
    geom_line(show.legend = FALSE) + 
    theme_bw() +
    ggtitle("Worldwide monthly new deaths") +
    theme(legend.position="bottom")

ggplotly(plot_deaths_monthly)

# TODO
# world map of cumulative cases
```



### COVID-19 vaccinations

Now, let's take a look at the evolution of vaccinations, which started in late 2020.  

```{r}

# get non-NA vaccination reports
df_vac_nona <- df_vac %>%
  # filter to keep only non-NA in people_vaccinated_per_hundred
  drop_na(people_vaccinated_per_hundred) %>%
  # group by month, year
  group_by(month, year, country) %>%
  # select first available value of each month
  filter(date == min(date))
  
# let's see how the people_vaccinated_per_hundred evolved (ratio (in percent) between population immunized and total population up to the date in the country)
plot_vac_percent <- df_vac_nona %>%
  # filter(country %in% c("USA", "China", "India")) %>% # to show only some countries
  ggplot(aes(x = date, y = people_vaccinated_per_hundred, color = country, group = country)) + 
    geom_line(show.legend = FALSE) + 
    theme_bw() +
    ggtitle("People vaccinated with respect to the country's total population (%)") +
    theme(legend.position="bottom")

ggplotly(plot_vac_percent) # generate interactive plot

# let's see how the people_vaccinated_per_hundred evolved (ratio (in percent) between vaccination number and total population up to the date in the country) 
plot_total_vac <- df_vac_nona %>%
  # filter(country %in% c("United States", "China", "India")) %>% # to show only some countries
  ggplot(aes(x = date, y = total_vaccinations_per_hundred, color = country, group = country)) + 
    geom_line(show.legend = FALSE) + 
    theme_bw() +
    ggtitle("Total number of vaccine doses administered per 100 people in the total population (%)") +
    theme(legend.position="bottom")

ggplotly(plot_total_vac) # generate interactive plot


```

**Observations:** 

* Vaccination rates start in late 2020 and increase throughout 2021 and 2022. Some countries show a more drastic increase. 
* A number of countries did not report their vaccination rates throughout the included time period (interrupted lines)
* Vaccination rates vary widely, with some countries having most of their population vaccinated (90-100%) while others still report close to 0% vaccinated up until recently. 
* The number of vaccines received in each country is also very variable, with most countries reporting between 0-2 vaccines administered per person, and 2 countries reporting over 3 vaccinations on average (Gibraltar and Cuba). 

**Ideas:**

* It would be interesting to plot current vaccination rates by income type (high vs low)? 
* Maybe also plot the change in vaccination levels (some countries vaccinated very fast compared to others - what explains that? Access to the vaccine, smaller total population, both?) and/or the total number of vaccine doses (some countries already gave more than 3 shots, while others barely have started to vaccinate) by some country features
* Group countries by their most current vaccination rate and plot number of cases or deaths (are the cases less for countries with higher vaccination rates?) 


```{r}

# more data cleaning
# based on the vaccination plots above, we notice that some countries seem to have large "gaps" in their reporting of vaccination rates.
# Let's take a look at a few of these to see what's going on:
# df_vac[df_vac$country == "Pitcairn",]
# df_vac_nona[df_vac_nona$country == "Pitcairn",] # they only report 4 times
# df_vac[df_vac$country == "Saint Helena",]
# df_vac_nona[df_vac_nona$country == "Saint Helena",] # they also only report 4 times
# df_vac[df_vac$country == "United Arab Emirates",]
# df_vac_nona[df_vac_nona$country == "United Arab Emirates",] # they also only report 4 times
# df_vac[df_vac$country == "Afghanistan",]
# looks like some small islands have a very small population (e.g. Pitcairn has 47 current residents and they reported their vaccinations only once for all these people)

# Solution: filter out countries for which 90% of their rows are NAs (these countries report vaccine rates consistently enough to keep).
# Alternative: we could filter out countries with less than 6 data points (months) in total - which means they reported vaccinations for less than 1 year in total. Given that vaccinations started in later 2020 (so about 16 months ago), this should still be a generous cutoff. 
df_vac_nona_clean <- df_vac %>%
  # first group by country and keep only countries that have at least 
  
  ########################### UNCOMMENT THIS TO FILTER OUT COUNTRIES THAT DID NOT PROVIDE ENOUGH VACCINATION REPORTS
  # group_by(country) %>%
  ## get an idea of percent NAs per country
  ## summarise(
  ##   n_na = sum(is.na(people_vaccinated_per_hundred)),
  ##   n_total = length(people_vaccinated_per_hundred),
  ##   pct_na = sum(is.na(people_vaccinated_per_hundred)) / length(people_vaccinated_per_hundred) * 100
  ## )
  ##  remove countries for which 90% of their rows are NAs (these countries report vaccine rates consistently enough to keep)
  # filter(sum(is.na(people_vaccinated_per_hundred)) / length(people_vaccinated_per_hundred) < 0.9) %>% # 
  # ungroup() %>%        
  ###########################

  # filter to keep only non-NA in people_vaccinated_per_hundred
  drop_na(people_vaccinated_per_hundred) %>%
  # group by month, year
  group_by(month, year, country) %>%
  # select first available value of each month
  filter(date == min(date))


  
# let's generate the same plots again to see how that looks:
plot_vac_percent <- df_vac_nona_clean %>%
  # filter(country %in% c("USA", "China", "India")) %>% # to show only some countries
  ggplot(aes(x = date, y = people_vaccinated_per_hundred, color = country, group = country)) + 
    geom_line(show.legend = FALSE) + 
    theme_bw() +
    ggtitle("People vaccinated with respect to the country's total population (%)") +
    theme(legend.position="bottom")

ggplotly(plot_vac_percent) # generate interactive plot

# let's plot again
plot_total_vac <- df_vac_nona_clean %>%
  # filter(country %in% c("United States", "China", "India")) %>% # to show only some countries
  ggplot(aes(x = date, y = total_vaccinations_per_hundred, color = country, group = country)) + 
    geom_line(show.legend = FALSE) + 
    theme_bw() +
    ggtitle("Total number of vaccine doses administered per 100 people in the total population (%)") +
    theme(legend.position="bottom")

ggplotly(plot_total_vac) # generate interactive plot

# TODO
# world map of total number of vaccinations
# world map of total number of vaccinations
```

  
After removing countries that did not consistently report their vaccination rates (e.g., countries with more than 50% of their rows being NAs), this looks a lot cleaner! BUT NOT DOING THIS FOR NOW.


Let's bin the latest reported vaccination rate for every country into 4 groups: 

* Less than 20% of the total population is vaccinated
* Between 20-50% is vaccinated
* Between 50-80% is vaccinated
* More than 80% is vaccinated

```{r}

# select the most recent reported vaccination percent for every country
box_pct_vac <- df_vac_nona_clean %>%
  group_by(country) %>%
  filter(date == max(date)) %>%
  # create new binning variable
  mutate(binned_people_vaccinated_per_hundred = case_when(people_vaccinated_per_hundred < 30 ~ "Under 30%", 
                                                          people_vaccinated_per_hundred >= 30 & people_vaccinated_per_hundred < 50 ~ "30-50%",
                                                          people_vaccinated_per_hundred >= 50 & people_vaccinated_per_hundred < 70 ~ "50-70%",
                                                          people_vaccinated_per_hundred >= 70 & people_vaccinated_per_hundred < 90 ~ "70-90%",
                                                          people_vaccinated_per_hundred >= 90 ~ "Above 90%"
                                                          )) %>%
  ggplot(aes(x = forcats::fct_reorder(binned_people_vaccinated_per_hundred, -people_vaccinated_per_hundred, .fun = median), y = people_vaccinated_per_hundred, 
             group = binned_people_vaccinated_per_hundred, fill = binned_people_vaccinated_per_hundred, color = binned_people_vaccinated_per_hundred)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(aes(text=sprintf("Country: %s<br> Rate: %s", country, people_vaccinated_per_hundred))) +
  labs(title = "Countries binned by their current vaccination rate",
       x = "Current vaccination rates (March 2022)",
       y = "Percent people vaccinated per 100 in total population",
       color = "",
       fill = "") +
  theme_bw()

ggplotly(box_pct_vac, tooltip = "text")

```


## Merging COVID-19 cases and vaccinations

Let's now merge the COVID-19 cases and vaccination datasets. 

```{r}

# merge the cases and vaccinations dataset by country and date keeping all the df_cases rows
df_merged <- merge(df_cases, 
                   df_vac_nona_clean[ , -which(names(df_vac_nona_clean) %in% c("year", "month_num","month"))], 
                   by = c("country", "date"), all.x = TRUE)

# check that the merge is correct
# df_merged[df_merged$country == "Argentina",]
# df_vac_nona_clean[df_vac_nona_clean$country == "Argentina",]
# yep all seems good (most dates do not have vaccination reports, only certain ones like for instance 2022-03-01)

# find dates corresponding to important vaccination milestones
df_merged <- df_merged %>%
  group_by(country) %>%
  mutate(vaccination_events = as.factor(case_when(people_vaccinated_per_hundred >= 90 ~ "vac_90pct",
                                        people_vaccinated_per_hundred >= 70 ~ "vac_70pct",
                                        people_vaccinated_per_hundred >= 50 ~ "vac_50pct",
                                        people_vaccinated_per_hundred >= 30 ~ "vac_30pct",
                                        people_vaccinated_per_hundred >= 0 & people_vaccinated_per_hundred < 10 ~ "vac_start",
                                        TRUE ~ "no")),
         vaccination_events = replace(vaccination_events, duplicated(vaccination_events), NA) # keep only first occurrence of an event
         )  

# check outputs
# df_merged[df_merged$country == "Sweden", c("date", "people_vaccinated_per_hundred", "vaccination_events")]

# Add percent vaccinated bins variable to main df
bins_df <- df_vac_nona_clean %>%
  group_by(country) %>%
  filter(date == max(date)) %>%
  # create new binning variable
  mutate(binned_people_vaccinated_per_hundred = case_when(people_vaccinated_per_hundred < 30 ~ "Under 30%", 
                                                          people_vaccinated_per_hundred >= 30 & people_vaccinated_per_hundred < 50 ~ "30-50%",
                                                          people_vaccinated_per_hundred >= 50 & people_vaccinated_per_hundred < 70 ~ "50-70%",
                                                          people_vaccinated_per_hundred >= 70 & people_vaccinated_per_hundred < 90 ~ "70-90%",
                                                          people_vaccinated_per_hundred >= 90 ~ "Above 90%"
                                                          )) %>%
  select(country, binned_people_vaccinated_per_hundred)

# merge with main
df_merged <- merge(df_merged, bins_df, by = c("country"), all.x = TRUE)

# checking that there are no countries with more than 10% vaccinated at their vac_start mark
df_merged %>%
  filter(vaccination_events == "vac_start" & people_vaccinated_per_hundred > 10) %>%
  select(country, vaccination_events, people_vaccinated_per_hundred)

```

### Plotting cases in relation to important vaccination milestones

#### Percent vaccinated per hundred

```{r}
# filter some countries
vac_events_sub <- df_merged %>%
  filter(country %in% c("USA", "Brazil", "Cuba", "Italy", "France", "Germany", "Japan", "India", "Australia", "South Africa", "Egypt", "Sudan")) 

# create vaccination start df
vac_start_df <- vac_events_sub %>%
  select(country, date, vaccination_events) %>%
  filter(vaccination_events == "vac_start")

# create vaccination 30% df
vac_30_df <- vac_events_sub %>%
  select(country, date, vaccination_events) %>%
  filter(vaccination_events == "vac_30pct")

# create vaccination 30% df
vac_50_df <- vac_events_sub %>%
  select(country, date, vaccination_events) %>%
  filter(vaccination_events == "vac_50pct")

# create vaccination 30% df
vac_70_df <- vac_events_sub %>%
  select(country, date, vaccination_events) %>%
  filter(vaccination_events == "vac_70pct")

# create vaccination 30% df
vac_90_df <- vac_events_sub %>%
  select(country, date, vaccination_events) %>%
  filter(vaccination_events == "vac_90pct")


# daily cases with weekly average number of new cases
plot_cases_weekly <- vac_events_sub %>%
  ggplot() + 
    geom_histogram(aes(x = date, y = daily_new_cases, color = country, group = country), stat = "identity", alpha = 0.5, binwidth = 0.1) +
    geom_line(aes(x = date, y = weekly_new_cases), color = "darkblue", show.legend = FALSE) + 
    facet_wrap(~country) +
    # this will generate some warnings when not all countries have all milestones, nothing to worry about
    geom_vline(data = vac_start_df, aes(xintercept = as.numeric(date), text = sprintf("Start date: %s", date)), linetype = 2, color = "black") +
    geom_vline(data = vac_30_df, aes(xintercept = as.numeric(date), text = sprintf("30 pct vaccinated: %s", date)), linetype = 2, color = "grey40") +
    geom_vline(data = vac_50_df, aes(xintercept = as.numeric(date), text = sprintf("50 pct vaccinated: %s", date)), linetype = 2, color = "grey50") +
    geom_vline(data = vac_70_df, aes(xintercept = as.numeric(date), text = sprintf("70 pct vaccinated: %s", date)), linetype = 2, color = "grey60") +
    geom_vline(data = vac_90_df, aes(xintercept = as.numeric(date), text = sprintf("90 pct vaccinated: %s", date)), linetype = 2, color = "grey70") +
    # geom_text(data = vac_start_df, aes(x = as.Date(date), y = 750000, label = date)) + # add static event labels
    theme_bw() +
    ggtitle("Worldwide daily new cases and vaccination milestones") +
    theme(legend.position="bottom", axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

ggplotly(plot_cases_weekly, tooltip = "text")

# and the number of new deaths?
plot_deaths_weekly <- vac_events_sub %>%
  ggplot() + 
    geom_histogram(aes(x = date, y = daily_new_deaths, color = country, group = country), stat = "identity", alpha = 0.5, binwidth = 0.1) +
    geom_line(aes(x = date, y = weekly_new_deaths), color = "darkblue", show.legend = FALSE) + 
    facet_wrap(~country) +
  # this will generate some warnings when not all countries have all milestones, nothing to worry about
    geom_vline(data = vac_start_df, aes(xintercept = as.numeric(date), text = sprintf("Start date: %s", date)), linetype = 2, color = "black") +
    geom_vline(data = vac_30_df, aes(xintercept = as.numeric(date), text = sprintf("30 pct vaccinated: %s", date)), linetype = 2, color = "grey40") +
    geom_vline(data = vac_50_df, aes(xintercept = as.numeric(date), text = sprintf("50 pct vaccinated: %s", date)), linetype = 2, color = "grey50") +
    geom_vline(data = vac_70_df, aes(xintercept = as.numeric(date), text = sprintf("70 pct vaccinated: %s", date)), linetype = 2, color = "grey60") +
    geom_vline(data = vac_90_df, aes(xintercept = as.numeric(date), text = sprintf("90 pct vaccinated: %s", date)), linetype = 2, color = "grey70") +
    # geom_text(data = vac_start_df, aes(x = as.Date(date), y = 750000, label = date)) + # add static event labels
    theme_bw() +
    ggtitle("Worldwide daily new deaths and vaccination milestones") +
    theme(legend.position="bottom", axis.text.x = element_text(angle = 45))

ggplotly(plot_deaths_weekly, tooltip = "text")
```

#### Total number of vaccine doses per hundred people

```{r}
# find dates corresponding to important vaccination milestones
df_merged <- df_merged %>%
  group_by(country) %>%
  mutate(vaccine_dose_events = as.factor(case_when(total_vaccinations_per_hundred >= 300 ~ "3_doses",
                                                   total_vaccinations_per_hundred >= 200 ~ "2_doses",
                                                   total_vaccinations_per_hundred >= 100 ~ "1_dose",
                                                   TRUE ~ "no")),
         vaccine_dose_events = replace(vaccine_dose_events, duplicated(vaccine_dose_events), NA) # keep only first occurrence of an event
         )  

# check outputs
# df_merged[df_merged$country == "Sweden", c("date", "total_vaccinations_per_hundred", "vaccine_dose_events")]
```


```{r}
# filter some countries
vac_events_sub <- df_merged %>%
  filter(country %in% c("USA", "Brazil", "Cuba", "Italy", "France", "Germany", "Japan", "India", "Australia", "South Africa", "Egypt", "Sudan")) 

# create 1st dose df
vac_1_dose <- vac_events_sub %>%
  select(country, date, vaccine_dose_events) %>%
  filter(vaccine_dose_events == "1_dose")

# create 2nd dose df
vac_2_doses <- vac_events_sub %>%
  select(country, date, vaccine_dose_events) %>%
  filter(vaccine_dose_events == "2_doses")

# create 3rd dose df
vac_3_doses <- vac_events_sub %>%
  select(country, date, vaccine_dose_events) %>%
  filter(vaccine_dose_events == "3_doses")

# daily cases with weekly average number of new cases
plot_cases_weekly <- vac_events_sub %>%
  ggplot() + 
    geom_histogram(aes(x = date, y = daily_new_cases, color = country, group = country), stat = "identity", alpha = 0.5, binwidth = 0.1) +
    geom_line(aes(x = date, y = weekly_new_cases), color = "darkblue", show.legend = FALSE) + 
    facet_wrap(~country) +
    # this will generate some warnings when not all countries have all milestones, nothing to worry about
    geom_vline(data = vac_1_dose, aes(xintercept = as.numeric(date), text = sprintf("1st dose: %s", date)), linetype = 2, color = "black") +
    geom_vline(data = vac_2_doses, aes(xintercept = as.numeric(date), text = sprintf("2nd dose: %s", date)), linetype = 2, color = "grey40") +
    geom_vline(data = vac_3_doses, aes(xintercept = as.numeric(date), text = sprintf("3rd dose: %s", date)), linetype = 2, color = "grey50") +
    # geom_text(data = vac_start_df, aes(x = as.Date(date), y = 750000, label = date)) + # add static event labels
    theme_bw() +
    ggtitle("Worldwide daily new cases and vaccine doses milestones") +
    theme(legend.position="bottom", axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

ggplotly(plot_cases_weekly, tooltip = "text")

# and the number of new deaths?
plot_deaths_weekly <- vac_events_sub %>%
  ggplot() + 
    geom_histogram(aes(x = date, y = daily_new_deaths, color = country, group = country), stat = "identity", alpha = 0.5, binwidth = 0.1) +
    geom_line(aes(x = date, y = weekly_new_deaths), color = "darkblue", show.legend = FALSE) + 
    facet_wrap(~country) +
  # this will generate some warnings when not all countries have all milestones, nothing to worry about
    geom_vline(data = vac_1_dose, aes(xintercept = as.numeric(date), text = sprintf("1st dose: %s", date)), linetype = 2, color = "black") +
    geom_vline(data = vac_2_doses, aes(xintercept = as.numeric(date), text = sprintf("2nd dose: %s", date)), linetype = 2, color = "grey40") +
    geom_vline(data = vac_3_doses, aes(xintercept = as.numeric(date), text = sprintf("3rd dose: %s", date)), linetype = 2, color = "grey50") +
    # geom_text(data = vac_start_df, aes(x = as.Date(date), y = 750000, label = date)) + # add static event labels
    theme_bw() +
    ggtitle("Worldwide daily new deaths and vaccine doses milestones") +
    theme(legend.position="bottom", axis.text.x = element_text(angle = 45))

ggplotly(plot_deaths_weekly, tooltip = "text")
```

### Covid case trends by vaccination group 


Here, we visualize COVID new cases/deaths trends separately for each vaccination group (more than 90% vaccinated, more than 70% vaccinated, etc.). 

```{r}

# daily cases with weekly average number of new cases
plot_cases_weekly <- df_merged %>%
  ggplot() + 
    geom_histogram(aes(x = date, y = daily_new_cases, 
                       color = binned_people_vaccinated_per_hundred, 
                       group = binned_people_vaccinated_per_hundred), stat = "identity", alpha = 0.5, binwidth = 0.1) +
    geom_line(aes(x = date, y = weekly_new_cases), color = "darkblue", show.legend = FALSE) + 
    facet_wrap(~binned_people_vaccinated_per_hundred) +
    theme_bw() +
    ggtitle("Worldwide daily new cases by vaccination rates") +
    theme(legend.position="bottom", axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

ggplotly(plot_cases_weekly, tooltip = "text")

# and the number of new deaths?
plot_deaths_weekly <- df_merged %>%
  ggplot() + 
    geom_histogram(aes(x = date, y = daily_new_deaths, 
                       color = binned_people_vaccinated_per_hundred,
                       group = binned_people_vaccinated_per_hundred), stat = "identity", alpha = 0.5, binwidth = 0.1) +
    geom_line(aes(x = date, y = weekly_new_deaths), color = "darkblue", show.legend = FALSE) + 
    facet_wrap(~binned_people_vaccinated_per_hundred) +
    theme_bw() +
    ggtitle("Worldwide daily new deaths and vaccine doses milestones") +
    theme(legend.position="bottom", axis.text.x = element_text(angle = 45))

ggplotly(plot_deaths_weekly, tooltip = "text")


```
This isn't really showing something very meaningful, may want to cut off and look at cases for instance once at least 50% of people have been vaccinated, or once most people have had the 1st dose. 


```{r}
# save merged df (this will be used to merge with the sustainability data)
write.csv(df_merged, "data/covid_cases_vaccines_clean.csv")
```

