---
title: "COVID_Sustainability_Global"
author:
- Diego G. Davila
- Margaret Gardner
- Joelle Bagautdinova
date: 'Due before midnight, May 1st'
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
  pdf_document:
    toc_depth: '4'
    number_sections: yes
urlcolor: blue
---

```{r Setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output

# Package setup
if(!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, dplyr, ggplot2, ggthemes, data.table, lubridate,
               GGally, RColorBrewer, ggsci, plotROC, usmap,
               plotly, ggpubr, vistime, skimr, glmnet, dgCMatrix, leaps, car)
```



Load outputs from EDA Rmds
```{r load.dfs}
load("data/WorldSustainabilityData_Processed.Rda")
#skim(wsd)
max(wsd$Year)
wsd <- wsd %>% dplyr::select(1:3,"Continent",4:21,23:ncol(wsd)) %>% # rearrange a little to group continent with the other geographic data
  mutate(Country.Name=as.factor(Country.Name)) %>%
  rename(country=Country.Name)

load("data/covid_cases_vaccines_clean.Rda")
#skim(covid)
```

```{r}
#some discrepancy in how countries are named, gonna fix that
wsd.country <- levels(wsd$country)
covid.country <- levels(covid$country)
#setdiff(wsd.country, covid.country) #only in wsd, not covid
#setdiff(covid.country, wsd.country) #only in covid, not wsd

covid$country <- recode_factor(covid$country, "UK"="United Kingdom", "Gambia"="Gambia, The", "Laos"="Lao PDR", "Egypt"="Egypt, Arab Rep.", "South Korea"="Korea, Rep.", "USA"="United States", "Russia"="Russian Federation", "Venezuela"="Venezuela, RB", "Democratic Republic Of The Congo"="DRC")
wsd$country <- recode_factor(wsd$country,  "Iran, Islamic Rep."="Iran", "Bosnia and Herzegovina"="Bosnia And Herzegovina", "Cote d'Ivoire"="Cote D Ivoire", "Guinea-Bissau"="Guinea Bissau", "Kyrgyz Republic"="Kyrgyzstan", "North Macedonia"="Macedonia", "Eswatini"="Swaziland", "Vietnam"="Viet Nam", "Timor-Leste"="Timor Leste", "Slovak Republic"="Slovakia", "Congo, Rep."="Congo", "Congo, Dem. Rep."="DRC")
```

Before attempting to merge, I'll parse down the dataframes so we can hopefully we can mix and match as needed to get what we need for our final analyses.

## WSD Summary Vales

To start, I'll make a wide df the WSD data into the COVID dataset containing: the latest (2018) data; average of each parameter over time (2001 to 2018); the average of the last 3 years of data (2016-2018 for most countries); change in each parameter over time (2001-2018). Also made one for latest date for each country since 32 are missing 2018 data, not sure which we want to use!

``` {r}
# 2018 values
wsd.2018 <- wsd %>% subset(Year==max(wsd$Year)) %>% select(-Year) #filter to last year of WSD data
colnames(wsd.2018)[4:30] <- paste(colnames(wsd.2018)[4:30], "2018", sep = "_") #add suffix
head(wsd.2018)
n_unique(wsd.2018$country) 
```

```{r}
# latest data for each country
wsd.latest <- wsd %>% arrange(Year) %>%
  group_by(country) %>%
  slice_tail(n=1) %>%
  ungroup()
colnames(wsd.latest)[2:31] <- paste(colnames(wsd.latest)[2:31], "latest", sep = "_") #add suffix
table(as.factor(wsd.latest$Year_latest)) # see yr frequencies
```

Values averaged across 2016-2018; only includes countries with da. 

```{r}
# 3-yr averages for numeric
wsd.3yr.num <- wsd %>% arrange(Year) %>%
  filter(Year >= 2016) %>% #get 2016-2018 data
  group_by(country) %>%
  filter(n() == 3) %>%
  summarise_if(is.numeric, mean) %>% #means
  ungroup()
colnames(wsd.3yr.num)[2:26] <- paste(colnames(wsd.3yr.num)[2:26], "3yr", sep = "_") #add suffix


# 3-yr mode for factors
# create a function to find the mode. Taken from: https://stackoverflow.com/questions/2547402/how-to-find-the-statistical-mode
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

wsd.3yr.fac <- wsd %>% arrange(Year) %>%
  group_by(country) %>%
  filter(Year >= 2016) %>% #get 2016-2018 data
  filter(n() == 3) %>%
  mutate(WorldBankIncomeClass = Mode(WorldBankIncomeClass),
         RegimeType = Mode(RegimeType),
         WorldRegion=Mode(WorldRegion),
         nYears=n()) %>% #count how many years are being averaged for each country) %>%
  ungroup() %>%
  select(country, nYears, WorldBankIncomeClass, RegimeType, WorldRegion) %>%
  distinct()
#n_unique(wsd.3yr.fac$country) #make sure no countries missing
colnames(wsd.3yr.fac)[2:5] <- paste(colnames(wsd.3yr.fac)[2:5], "3yr", sep = "_") #add suffix

wsd.3yr <- merge(wsd.3yr.fac, wsd.3yr.num, by="country") %>% #rejoin num and fac
    select(-Year_3yr, -nYears_3yr)
nrow(wsd.3yr)
```
Averaging across all years for every country with available data. 

```{r}
# averages for numeric
wsd.avg.num <- wsd %>% arrange(Year) %>%
  group_by(country) %>%
  summarise_if(is.numeric, mean) %>%
  ungroup() %>%
  select(-Year)
colnames(wsd.avg.num)[2:25] <- paste(colnames(wsd.avg.num)[2:25], "avg", sep = "_") #add suffix

# mode for factors
wsd.avg.fac <- wsd %>% arrange(Year) %>%
  group_by(country) %>%
  mutate(WorldBankIncomeClass = Mode(WorldBankIncomeClass),
         RegimeType = Mode(RegimeType),
         WorldRegion=Mode(WorldRegion),
         nYears=n()) %>% #count how many years are being averaged for each country
  ungroup() %>%
  select(country, WorldBankIncomeClass, RegimeType, WorldRegion, nYears) %>%
  distinct()
colnames(wsd.avg.fac)[2:5] <- paste(colnames(wsd.avg.fac)[2:5], "avg", sep = "_") #add suffix

# #make sure modes returned as expected 
# angola <- wsd$WorldBankIncomeClass[wsd$country=="Angola"]
# Mode(angola)
# wsd.avg.fac$WorldBankIncomeClass[wsd.avg.fac$country=="Angola"]
# #looks good!

wsd.avg <- merge(wsd.avg.fac, wsd.avg.num, by="country") #rejoin num and fac
table(as.factor(wsd.avg$nYears_avg)) # see yr frequencies
```

As you can see from the frequency table above, only 60 countries have data for all 18 years. We could restrict this to only including the averages of countries with at least 15 yrs of data by uncommenting the chunk below (filters df down to 99 countries):

```{r}
# #uncomment to remove any countries that have less than 15 years worth of data
# wsd.avg <- wsd.avg %>%
#   subset(nYears_avg >= 15)
# nrow(wsd.avg)
```


Change in each factor over 10 years (2008-2018)

```{r}
# change in each parameter over 10 years

# define function to get relative change
rel.diff <- function(x) {(x - lag(x))/lag(x)} # define function to get relative change


wsd.change.num <- wsd %>% 
  group_by(country) %>%
  arrange(Year) %>%
  select_if(is.numeric)%>% # just dealing with numeric for now
  subset(Year==2018 | Year == 2008) %>%
  filter(n() >= 2) %>% #make sure country has data for both years
  mutate(across(where(is.numeric), ~rel.diff(.), .names="{.col}_rel.diff")) %>% # change normalized by baseline values
  mutate(across(c(!ends_with("diff")), ~diff(.x), .names="{.col}_diff"))  %>%  #absolute change 
  mutate_if(is.numeric, function(x) ifelse(is.infinite(x), 0, x)) %>% # change Inf values to 0
  select(-Year, -Year_diff, -Year_rel.diff) %>%
  drop_na() %>%
  ungroup()

#now factors
wsd.change.fac <- wsd %>%
  select(country, Year, WorldBankIncomeClass, RegimeType)
# order factors
wsd.change.fac$WorldBankIncomeClass <- factor(wsd.change.fac$WorldBankIncomeClass, levels=c("Low income", "Lower-middle income", "Upper-middle income", "High income"), ordered=TRUE)
wsd.change.fac$RegimeType <- factor(wsd.change.fac$RegimeType, levels=c("Closed Autocracy", "Electoral Autocracy", "Electoral Democracy", "Liberal Democracy"), ordered=TRUE)

wsd.change.fac <- wsd.change.fac %>% 
  group_by(country) %>%
  arrange(Year) %>%
  subset(Year==2018 | Year == 2008) %>%
  filter(n() >= 2) %>% #make sure country has data for both years
  mutate(WorldBankIncomeClass_diff = unclass(WorldBankIncomeClass) - unclass(lag(WorldBankIncomeClass)),
         WorldBankIncomeClass_rel.diff = rel.diff(unclass(WorldBankIncomeClass)),
         RegimeType_diff = unclass(RegimeType) - unclass(lag(RegimeType)),
         RegimeType_rel.diff = rel.diff(unclass(RegimeType))) %>% 
  mutate_if(is.numeric, function(x) ifelse(is.infinite(x), 0, x)) %>% # change Inf values to 0
  select(-Year) %>%
  drop_na() %>%
  ungroup()

#now join and drop unnecessary columns used for calculations
wsd.change <- merge(wsd.change.num, wsd.change.fac, by ="country") %>%
  select(country, ends_with("diff"))
```

Merge into one wide wsd dataframe:

```{r}
# one predictor dataframe to rule them all
wsd.df_list <- list(wsd.2018, wsd.3yr, wsd.avg, wsd.change) #dfs to merge
wsd.sum.wide <- wsd.df_list %>% reduce(full_join, by='country') #join
#skim(wsd.sum.wide)

#here's a long version if you want it
wsd.sum.long <- wsd.sum.wide %>%
  select(-WorldBankIncomeClass_rel.diff, -RegimeType_rel.diff) %>%
  mutate(WorldBankIncomeClass_diff=as.factor(WorldBankIncomeClass_diff),
         RegimeType_diff=as.factor(RegimeType_diff)) %>%
  pivot_longer(c(contains("_")), 
    names_to = c(".value", "type"), 
    names_sep = "_", 
    values_drop_na = TRUE
  )

head(wsd.sum.wide)
```

## COVID Summary Values

I'll attempt to make the dataframe smaller and more mergable by filtering down into summary variables, similar to WSD. Some possibilities of things we could compare between countries:
- Date vaccinations start
- mean daily vaccinations during first 6 months of availability
- peak daily new deaths & date of that peak (actual date or number days from start)
- days to binned_people_vaccinated_per_hundred >= X
- binned_people_vaccinated_per_hundred as of March 2022
- percent vaccines as of March 2022

```{r}
skim(covid)
#last day with data for each country
last.timepoint <- covid %>% group_by(country) %>%
  arrange(country,date) %>%
  slice_tail(n=1) %>%
  ungroup()
last.timepoint <- last.timepoint[, which(colMeans(!is.na(last.timepoint)) > 0.5)] #remove cols with NA > 50%
skim(last.timepoint)

# get missing column names
missing_vac_cols <-  names(which(colMeans(is.na(last.timepoint)) > 0.5))

# add last reported (within 2022) vaccination info
vac_info_2022 <- covid %>%
  drop_na(people_vaccinated_per_hundred) %>%   # filter to keep only non-NA in people_vaccinated_per_hundred
  group_by(country) %>% 
  filter(date == max(date)) %>%   # keep last reported date for every country
  filter(year == 2022) %>% # keep only countries who last reported in 2022 (as we are interested in a recent vaccination status)
  # filter(month_year == "2022-03-01") # could also filter for march 2022 specifically
  select(country, date, missing_vac_cols) %>% # keep only variables not yet included in last.timepoint
  rename(date.vac = date)# rename the date to correspond to last reported vaccine info

# merge the last reported vaccination info with last reported cases/deaths info
last.timepoint <- merge(last.timepoint, vac_info_2022, by = "country")
colnames(last.timepoint)[2:28] <- paste(colnames(last.timepoint)[2:28], "last", sep = "_") #add suffix


# #find date of vaccination events
# table(covid$vaccination_events)
# table(covid$vaccine_dose_events)
# table(covid$binned_people_vaccinated_per_hundred)
```

Vaccine outcomes:

```{r}
# creating dfs of vaccination events - much of this is from covid_EDA.Rmd and can be removed/cleaned up once Rmds are merged!

vac_start_df <- covid %>%
  select(country, date, vaccination_events, people_vaccinated_per_hundred, people_vaccinated) %>%
  filter(vaccination_events == "vac_start") %>%
  mutate(days.to.start_rel = as.numeric(difftime(date, min(date), units = "days")),
         days.to.start_abs = as.numeric(difftime(date, "2020-12-01", units = "days"))) %>%
  rename(date.first.vac = date)

# adding date of first vaccination for each country to the covid df so it can be used as a baseline later -
vac_start_date <- vac_start_df %>% select(country, date.first.vac)
covid <- full_join(covid, vac_start_date, by ="country")

vac_1_dose <- covid %>%
  select(country, date, vaccine_dose_events, date.first.vac) %>% #
  filter(vaccine_dose_events == "1_dose")%>%
  mutate(days.to.1d_rel = as.numeric(difftime(date, min(date), units = "days")),
         days.to.1d_abs = as.numeric(difftime(date, "2020-12-01", units = "days")), # setting Dec 1, 2020 as vaccine baseline to compare across vaccine events
         days.to.1d_fromstart = as.numeric(difftime(date, date.first.vac, units = "days"))
         ) 
vac_1_dose[which.min(vac_1_dose$date),] #first country to get one dose per person

vac_2_doses <- covid %>%
  select(country, date, vaccine_dose_events, date.first.vac) %>%
  filter(vaccine_dose_events == "2_doses")%>%
  mutate(days.to.2d_rel = as.numeric(difftime(date, min(date), units = "days")),
         days.to.2d_abs = as.numeric(difftime(date, "2020-12-01", units = "days")),
         days.to.2d_fromstart = as.numeric(difftime(date, date.first.vac, units = "days"))
         )
vac_2_doses[which.min(vac_2_doses$date),] #first country to hit 2 doses

vac_50_df <- covid %>%
  select(country, date, vaccination_events, date.first.vac) %>%
  filter(vaccination_events == "vac_50pct") %>%
  mutate(days.to.50pct_rel = as.numeric(difftime(date, min(date), units = "days")),
         days.to.50pct_abs = as.numeric(difftime(date, "2020-12-01", units = "days")),
         days.to.50pct_fromstart = as.numeric(difftime(date, date.first.vac, units = "days"))
         )

# #percent of ppl w/ 1 dose of vaccine as of mar 17 2022 - redundant to last.timepoint
# binned_vac <- covid %>%
#   select(country, date, binned_people_vaccinated_per_hundred) %>%
#   drop_na(binned_people_vaccinated_per_hundred) %>%
#   group_by(country) %>%
#   arrange(date) %>%
#   slice_tail(n=1) %>%
#   ungroup() %>%
#   rename (bin.ppl.vac.per100_17mar22=binned_people_vaccinated_per_hundred)
#summary(binned_vac$date)
#n_unique(binned_vac$country) #no countries lost

vac_rate <- covid %>%
  select(country, date, daily_vaccinations_per_million, date.first.vac) %>% 
  filter(daily_vaccinations_per_million > 0,
         date <= (date.first.vac %m+% months(6))) %>% # daily vaccinations in 1st 6 mo - make relative to their own vaccine start date
  drop_na(daily_vaccinations_per_million) %>% 
  group_by(country) %>%
  summarise(daily.vac.per.mil_6moavg = mean(daily_vaccinations_per_million))

vac_maxrate <- covid %>%
  select(country, date, daily_vaccinations_per_million, daily_vaccinations_raw, daily_vaccinations) %>% 
  group_by(country) %>%
  mutate_if(is.numeric, ~replace(., is.na(.), -1)) %>% #temporarily replace na's with -1
  summarise_if(is.numeric, max) %>%
  na_if(-1) %>% #set -1's back to na
  rename(daily.vac.raw_max = daily_vaccinations_raw,
         daily.vac_max = daily_vaccinations,
         daily.vac.per.mil_max = daily_vaccinations_per_million)
```

``` {r}
# cases & deaths up until vaccines start
covid_prevac <- covid %>%
  group_by(country) %>%
  drop_na(date.first.vac) %>%
  subset(date < date.first.vac) %>%
  arrange(country,date) %>%
  slice_tail(n=1) %>%
  ungroup() %>%
  select("country"| contains("case") | contains("death"))
colnames(covid_prevac)[2:19] <- paste(colnames(covid_prevac)[2:19], "prevac", sep = "_") #add suffix


# cases & deaths at 1 yr after date.first.vac
covid_1yr <- covid %>%
  group_by(country) %>%
  drop_na(date.first.vac) %>%
  subset(date == (date.first.vac %m+% months(12))) %>%
  select("country"| contains("case") | contains("death"))
colnames(covid_1yr)[2:19] <- paste(colnames(covid_1yr)[2:19], "1yr", sep = "_") #add suffix
#covid_1yr <- covid_1yr[, which(colMeans(!is.na(covid_1yr)) > 0.5)] #remove cols with NA > 50%
#sum(is.na(covid_1yr$cumulative_total_cases))

#peak rates up until vaccinations start
covid_maxrate <- covid %>%
  drop_na(date.first.vac) %>%
  subset(date < date.first.vac) %>% 
  select("country" | contains("new") | contains("daily")) %>%
  group_by(country) %>%
  mutate_if(is.numeric, ~replace(., is.na(.), -1)) %>% #temporarily replace na's with -1
  summarise_if(is.numeric, max) %>%
  na_if(-1) %>% #set -1's back to na
  rename(daily.new.cases.per.100_max = daily_new_cases_per_hundred,
         daily.new.deaths.per.100_max = daily_new_deaths_per_hundred,
         monthly.new.cases.per.100_max = monthly_new_cases_per_hundred,
         monthly.new.deaths.per.100_max = monthly_new_deaths_per_hundred,
         weekly.new.cases.per.100_max = weekly_new_cases_per_hundred,
         weekly.new.deaths.per.100_max = weekly_new_deaths_per_hundred) #using . and _ separators to match WSD in case we want to pivot long later
#n_unique(covid_maxrate$country) #one country lost filtering to pre "2021-02-20"

```


Now let's merge all these vaccination dfs into one dataframe we can try to predict - RUN PCA HERE!

``` {r}
#df for vaccine info
vac.df_list <- list(vac_1_dose, vac_2_doses, vac_start_df, vac_50_df, vac_rate, vac_maxrate)
#drop date and event info since this is contained in col names
for (i in 1:length(vac.df_list)) {
  vac.df_list[[i]] <- vac.df_list[[i]] %>% select(-c(contains("date"), contains("events")))
}

vac.sum.wide <- vac.df_list %>% reduce(full_join, by='country') %>% as.data.frame() #join
write.csv(vac.sum.wide, "data/vaccination_summary_wide.csv", row.names = FALSE)
```

Merging case & death dfs

```{r}
#df for cases and deaths
case.df_list <- list(covid_prevac, covid_1yr, covid_maxrate)
case.sum.wide <- case.df_list %>% reduce(full_join, by='country') %>% as.data.frame() #join
skim(case.sum.wide)
```

Merge all COVID summary data

```{r}
covid.df_list <- list(case.sum.wide, vac.sum.wide, last.timepoint)
covid.all <- covid.df_list %>% reduce(full_join, by='country') %>% as.data.frame() #join
n_unique(covid.all$country)
skim(covid.all)
```


## Merge! Merge! Merge!

``` {r}
#merging WSD predictors with vaccination summary data
vac.wsd.wide <- inner_join(vac.sum.wide, wsd.sum.wide, by="country")
#skim(vac.wsd.wide)

#merging WSD predictors with cases, deaths & maxrate data
case.wsd.wide <- inner_join(case.sum.wide, wsd.sum.wide, by="country")
#skim(case.wsd.wide)

#merging WSD predictors with latest COVID data (~March 17, 2022)
mar22.wsd.wide <- inner_join(last.timepoint, wsd.sum.wide, by="country")
#skim(mar22.wsd.wide)

#merging everything!
wsd.covid.all <- inner_join(covid.all, wsd.sum.wide, by="country")
skim(wsd.covid.all)
```

Any of the above can be recombined/merged further or pivoted long.

predict:
days to vaccine access
days from vaccine access to 50% ppl
% ppl vaccinated at march 2022 (absolute counts and binned)
doses administered at march 2022 (absolute counts and binned)





