---
title: "Linear Regression and Random Forest"
output: html_notebook
---

Setup
```{r}
library(pacman)
pacman::p_load(tree, rpart, randomForest, ranger, rattle, pROC, partykit, ggplot2, glmnet,lda, data.table, ISLR, car, dplyr, tidyverse, stats, skimr)
```

## Data Preparation

Read in the Data
```{r}
data.full <- read.csv('data/case.wsd.wide-predictors.csv')
head(data.full)
names(data.full)
```

merge with vaccination success index
```{r}
#read in
new.data.with.VSI <- read.csv("data/VaccinationSucessIndexData-New.csv")
yval <- new.data.with.VSI %>% dplyr::select("country", "VSI")
data.full <- inner_join(data.full, yval,  by=("country"))
skim(data.full)
```

We'll remove the variables that have a high missing rate, and then remove the countries that have NAs below that threshold
```{r}
data.90 <- data.full[, apply(data.full, 2, function(col)sum(is.na(col))/length(col)) < 0.10]
data.complete <- na.omit(data.90)
nrow(data.complete)
```

## Linear Regression

We will now try some linear models. Here we will explore the relationship between measures capturing the quality of a nation's economy over the last reported 18 years and vaccination success. 

Variables:
* `ExportGoodsServices.GDP_avg`
* `FinalConsumptionExpenditure.GDP_avg`
* `GDP.Current_avg`
* `GDP.PerCapita.Current_avg`
* `ImportGoodsServices.GDP_avg`
* `ConsumerPriceInflation_avg`


```{r}

linear.fit.1 <- lm(data = data.complete, formula = VSI ~ ExportGoodsServices.GDP_avg + FinalConsumptionExpenditure.GDP_avg + GDP.Current_avg + GDP.PerCapita.Current_avg +
                     ImportGoodsServices.GDP_avg +
                     ConsumerPriceInflation_avg)
summary(linear.fit.1)

```

Fine tune the model, removing non-significant variables. 
```{r}

linear.fit.1.refined <- lm(data = data.complete, formula = VSI ~ GDP.PerCapita.Current_avg + ConsumerPriceInflation_avg)
summary(linear.fit.1.refined)
```

The GDP per capita and consumer price inflation are significant as predictors of the Vaccination Success Index at the 0.05 level. This indicates that an increase in the average per-capita GDP of a nation, and decrease in the average consumer price inflation, over the last reported 18 years, correspond to an increase in the success of that country's vaccination campaign. 

## Random Forest for Regression


Let us tune the parameters mtry (number of variables to sample) and ntree (number of trees to use). 
```{r}
set.seed(1) # for reproducibility, we set the seed
fit.1.rf <- randomForest(VSI~., data.complete, mtry=6, ntree=500)
plot(fit.1.rf)
```

Based on out of bag error, it seems 100 trees is sufficient. Now, let's tune mtry. 

```{r}
error.p <- 1:150  
for (p in 1:150)  
{
  rf.temp <- randomForest(VSI~., data.complete, mtry=p, ntree=100)
  error.p[p] <- rf.temp$mse[100]  
}
plot(1:150, error.p, pch=16,
     main = "Testing errors of mtry with 100 trees",
     xlab="mtry",
     ylab="OOB mse of mtry")
lines(1:150, error.p)

```

It seems that an mtry between 50 and 70 is appropriate. Accordingly, we will select a final mtry of 60. 

Now, let's develop the final model
```{r}
fit.final.rf <- randomForest(VSI~., data.complete, mtry=60, ntree=100)
plot(fit.final.rf)
```







