---
title: "Vaccination Success PCA"
output: html_notebook
---

Load packaged
```{r}
library(pacman)
p_load(dplyr, tidyverse, ggplot2, car, skimr, nnet, randomForest, factoextra)
```

Read Summary Variables
```{r}
vax.summary <- read.csv('data/vaccination_summary_wide.csv')
skim(vax.summary)
```

That's  lot of missing values... let me extract just the columns that have at least 70%. Then, remove the rows with NAs. 
```{r}
vax.70 <- vax.summary[, apply(vax.summary, 2, function(col)sum(is.na(col))/length(col)) < 0.30]
skim(vax.70)
vax.clean <- na.omit(vax.70)
skim(vax.clean)
```

Ok, now let's apply PCA on the cleaned dataset to see if we can develop some kind of vaccination success index. 
```{r}
vax.pca <- prcomp(x = vax.clean[,-1], scale. = TRUE, center = TRUE)
vax.pca$rotation
summary(vax.pca)
```

Percent Variance Explained 
```{r}
plot(summary(vax.pca)$importance[3,], ylab = "Cummulative Pct. Variance Explained", xlab = "PC", main = "Cummulative PVE")
```

Ok, let's see if we can trim some fat. Seems not all the summary variables are especially useful. 
```{r}
vax.pca2 <- prcomp(x = vax.clean[,c(-1, -3, -5, -7, -8)], scale. = TRUE, center = TRUE)
vax.pca2$rotation
summary(vax.pca2)
```

Looks like PC1 can be interpreted as how poorly a country did (take the negative sign to indicate how well it did). 

append it to the csv and save
```{r}
data.with.VSI <- vax.clean # copy the data
data.with.VSI$VSI <- -vax.pca2$x[,1] # append the VSI (inverse signed PC1 scores)

write.csv(data.with.VSI, "data/VaccinationSucessIndexData.csv", row.names = FALSE)


```


Rerunning with alternate df

``` {r}
vax.summary.new <- read.csv('data/vaccination_summary.csv')
vax.summary.new <- na.omit(vax.summary.new)
nrow(vax.summary.new)
vax.pca3 <- prcomp(x = vax.summary.new[,-1], scale. = TRUE, center = TRUE)
vax.pca3$rotation
summary(vax.pca3) # adding 2 more vars brings the proportion of variance of PC1 from 67 to 73%, so keeping those in here
plot(summary(vax.pca3)$importance[2, ],  # PVE
     ylab="PVE",
     xlab="Number of PC's",
     pch = 16, 
     main="Scree Plot of PVE for Vaccination variables")
# Following the elbow rule, the first PC1 clearly explains a sufficient amount of variance. We will thus keep the first PC. 

new.data.with.VSI <- vax.summary.new # copy the data
new.data.with.VSI$VSI <- -vax.pca3$x[,1] # append the VSI (inverse signed PC1 scores), where a positive nb means a country was good at getting people vaccinated

write.csv(new.data.with.VSI, "data/VaccinationSucessIndexData-New.csv", row.names = FALSE)
```

