---
title: "World Sustainability Dataset EDA + Plotting"
group: "Group 18: Diego G. Davila Margaret Gardner, Joelle Bagautdinova"
output: html_notebook
---

# World Sustainability Dataset EDA
This dataset was aggregated from several sources initially for the TrueCue Women+Data Hackathon, and then made available to the larger public. Dataset can be found [here](https://www.kaggle.com/datasets/truecue/worldsustainabilitydataset). This dataset tracks the performance of 173 counties in terms of several measures of global sustainability over 19 years. 

```{r setup, echo=FALSE}
library(pacman)
p_load(dplyr, tidyverse, ggplot2, car, skimr, nnet, randomForest)
```

## Processing

Read in data and do initial exploration
```{r}
data.raw <- read.csv('data/WorldSustainabilityDataset.csv', header = TRUE)
head(data.raw)
```

Looks like there are some missing elements labeled as empty instead of NA. Let's fix this before proceeding.
```{r}
data.raw[data.raw == ""] <- NA
```

There are quite a few missing fields. Let's get fields have at least a 90% complete rate. 
```{r}
data.90 <- data.raw[, apply(data.raw, 2, function(col)sum(is.na(col))/length(col)) < 0.10]
skim(data.90)
```

What do we get if we remove the NAs. Hoe many countries do we still have data on?
```{r}
data.processed <- na.omit(data.90)
# How many countries?
n.countries <- length(unique(data.processed$Country.Name))
n.countries
```

Ok good, now let's rename some of our variables to be easier to read and work with
```{r}
names(data.processed)[4] <- "Electricity.Access"
names(data.processed)[5] <- "Adj.SavingNetCO2Damage"
names(data.processed)[6] <- "Adj.SavingNaturalResourceDepletion"
names(data.processed)[7] <- "Adj.SavingNetForestDepletion"
names(data.processed)[8] <- "Adj.SavingParticulateEmissions"
names(data.processed)[9] <- "CompulsoryEducationDurationYears"
names(data.processed)[10] <- "ExportGoodsServices.GDP"
names(data.processed)[11] <- "FinalConsumptionExpenditure.GDP"
names(data.processed)[12] <- "GDP.Current"
names(data.processed)[13] <- "GDP.PerCapita.Current"
names(data.processed)[14] <- "Govt.FinalConsumptionExpenditure.GDP"
names(data.processed)[15] <- "GrossNationalExpenditure.GDP"
names(data.processed)[16] <- "ImportGoodsServices.GDP"
names(data.processed)[17] <- "ConsumerPriceInflation"
names(data.processed)[18] <- "ProportionParliamentSeats.Women"
names(data.processed)[19] <- "RenewableEnergyConsumption.pctOfTotal"
names(data.processed)[20] <- "Trade.GDP"
names(data.processed)[21] <- "WomenInBusinessLawIndex"
names(data.processed)[22] <- "AnnualProductionBasedCO2Emissions"
names(data.processed)[24] <- "WorldBankIncomeClass"
names(data.processed)[25] <- "IndividualsUsingInternet"
names(data.processed)[26] <- "LifeExpenctancy"
names(data.processed)[27] <- "TotalPopulation"
names(data.processed)[28] <- "RegimeType"
names(data.processed)[29] <- "RuralPopulation.Prop"
names(data.processed)[30] <- "TotalNaturalResources.GDP"
names(data.processed)[31] <- "UrbanPopulation.Prop"
names(data.processed)[32] <- "WorldRegion"
names(data.processed)
```
Set Certain Things to Factor
```{r}
data.processed$RegimeType <- as.factor(data.processed$RegimeType)
data.processed$WorldRegion <- as.factor(data.processed$WorldRegion)
data.processed$WorldBankIncomeClass <- as.factor(data.processed$WorldBankIncomeClass)

```


Export the processed data
```{r, eval=FALSE}
write.csv(data.processed, "data/WorldSustainabilityData_Processed.csv", row.names = FALSE)
```

## EDA

Now that we have processed our data, let's do some EDA

First let's plot some measures by country. 

Prevalence of Regime Types
```{r}

ggplot(data.processed, aes(x=fct_infreq(RegimeType), fill=fct_infreq(RegimeType))) +
  geom_bar(show.legend = FALSE)

```


